The Key to a Data Parallel Compiler

### 2016-04-05 2016-spring-array/2016-array-key.tex
Fix up formatting on a lot of lines and get things mostly working.

There has been some interesting use of GPUs to accelerate certain
compiler analyses and passes.  Unfortunately, attempts to parallelize
compilers usually takes the form of a great deal of work spent
to highly optimize a single compiler pass. The results are often
fragile and require a great deal of effort to maintain and keep in
line with the rest of the compiler. This integration problem makes it
difficult to scale the "ad hoc" parallelization effort of individual
compiler passes to a complete system. This is particularly true in
situations where the compiler passes may change rapidly, such as with
XSLT or document publishing, where it is very common for the document
preparer to write or tweak transformation passes to suit the specific
class of documents being handled. In these cases, the expertise and
time involved in parallelizing such transformations makes the ad hoc
approach mostly intractable.

Furthermore, compilers are increasingly becoming a part of a "live"
and interactive feedback loop.  Just in time compilation and on
the fly static compilation are both used to provide custom binaries
tuned to specific tasks on the fly. These techniques are used in web
services and large scale databases, where in-memory databases may
construct functions on the fly to execute. There is great interest
in moving such systems on to the GPU or other highly data-parallel
architectures, but it would significantly hamper the performance of
the system if the compilation phase of query execution were required
to occur on the CPU, especially at scale where there may be many more
GPU accelerators than CPU resources.

@@

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{\^}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives}
\label{tab:scalarprims}
\end{table}

@@

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Monadic/Unary Operators, each takes a single left
operand and describes a verb operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a verb along cells of an array \\
\end{tabular}
\caption{Primitive Dyadic/Binary Operators, each takes a left
 and right operand and describes a verb operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

@@

The Key operator (written \verb;⌸;), takes a single function
that expects two arguments and returns a function which takes two
arguments. The left argument is a set of keys, and the right argument
is a set of corresponding elements associated with those keys. In our
case, we provide matrices for both of these arguments, so each pair
of rows corresponds to a key-value pair. A simple example of the key
operation at work is to compute a histogram (≢ here can be thought
of as “tally” or count):

To understand a bit better how the Key operator applies its function,
consider the function {⍺ ⍵} which returns the pair of its right
and left arguments. If we apply it to the same value as above, we
get the following:

In our case, we use either Fk or Ek as our keys applied to the
corresponding AST. We also will drop off the first row in each AST
using 1↓ since this node “contains” everything. In a complete
AST this is usually the Module boundary node which contains the entire
set of functions and values in the module.

@@

We represent the AST as a matrix of 3 columns and n rows, one for
each node in the tree. The first column contains the inter-node
relationships in the form of a depth vector. The second column is a
vector of the node types, while the third contains the “value”
of the node, such as the name in a variable.

@@

The depth vector for these trees we name Fd and Ed, respectively:

@@

The node types we call Ft and Et:

@@

And finally, we call the values vectors Fv and Ev:

@@

We combine these to form the respective AST matrices. We write A,B
to catenate arrays A and B along their last axes and ⍪A to turn a
vector into a 1-column matrix. Thus, the two tree matrices are given
by the following expressions:

\begin{verbatim}
      Fd,Ft,⍪Fv
0 F f
1 E 0
2 F 0
3 E 0
4 V ⍵
4 P +
4 V ⍵
2 A 0
3 N 7
      Ed,Et,Ev
0 E v
1 E 0
2 V a
2 P +
2 V b
1 P ÷
1 E 0
2 V c
2 P ×
2 V d
\end{verbatim}

@@

\begin{enumerate}[noitemsep]
\item Are the nodes the same?
\item Are they siblings?
\item Does one appear “earlier” in the tree?
\item Are they at the same depth?
\item Is one an ancestor of another?
\end{enumerate}

@@

We write f⌿A to reduce the first axis of A using function f. Thus,
+⌿V is the sum of the elements in vector V. The function x⌈y
gives the maximum of its two arguments. So, the depth of each tree
is given by the following:

@@

The function table or outer product of f over vectors U and V is
written U ∘.f V giving a U V shaped matrix as a result. Thus,
(⍳3)∘.×⍳3 gives a small multiplication table:

@@

If we use = instead, we have a Boolean identity matrix:

@@

If we use ∘.= on the depth vector and its set of depths instead,
we see an expanded Boolean representation of the depth vector:

@@

These matrices let us see the nesting features of each tree more
visually, but also suggest another step. We can compute a sum scan
with +⍀, also called a prefix sum, along the first axis. Applying
this function on the above matrices leads to an interesting result:

@@

The expression V f⍤¯1⊢M applies f to corresponding elements of
V and rows of M:

@@

If n↑V takes the first n elements of V, then we can obtain coordinate
matrices from the prefix sums by noting that the spurious digits
all come after column d+1 where d is the depth corresponding to that
coordinate. The following gives a complete expression for computing
a node coordinate matrix from a depth vector, shown using Fd and Fe:

@@

The simplest operation over a node coordinate is to extract the depth
of the node. The expression C⍳0 finds the first occurrence of 0 in C
and returns the index of that occurrence. (The application of ⍳ here
is its dyadic form, meaning that it receives two arguments. Previous
uses of ⍳ have used its monadic form, which generates all indices,
instead of finding a specific index. Most APL functions have a
dyadic and monadic form.)  Thus, the depth of a node is given by
the following:

@@

We write (f g h) to represent the composition of functions f, g,
and h as a function train. This i s used with another shorthand of
composition according to the following equivalences:

\begin{verbatim}
A(f g h)B ←→ (A f B) g (A h B)
A(0 f h)B ←→ 0 f (A h B) ⍝ 0 is a constant
A(f g)B   ←→ f (A g B)
\end{verbatim}

@@

Here P is the coordinate we wish to check against C, to determine
whether C is a descendant. The functions are all standard logical
functions extended to apply point-wise over arrays. The function
⊢ always returns its right argument. The function (=∨0=⊢)
reads intuitively as, “equal or right element is zero.” In this
case to make this a predicate we combine this with the “for all”
reduction using ∧⌿. This particular pattern is a special case
of inner product, which we can compute using f.g. Thus, +.× is a
function that computes traditional matrix-matrix multiplication when
applied to two matrices. However, we can also use it to compute the
above reduction and application:

@@

This extends the reduction to allow us to use matrix values for
C and P instead of vectors, and thus determine whether rows in C
are descendants of multiple candidates given by P. This becomes
important to using the prefix function. In our examples, we want
to determine the nearest ancestor of a certain node type to which a
node belongs. In the case of function lifting, we want to determine
the nearest ancestor function, in the case of expression flattening,
we want to determine the nearest ancestor expression node. We can
extract the rows we care about for each example using ⌷, where i⌷M
returns the i-th row of M. In the following expressions, we use 0 2
rather than ¯1 in ⍤ to indicate that ⌷ will receive the entire
matrix on the right for each scalar index on the left.

@@

The matrices Fp and Ep correspond to the function and expression
node coordinates in our two examples, respectively. We could now
use Fp and Ep to compare against Fc and Ec and determine which nodes
contained which nodes in the tree. However, our prefix function will
return 1 when we ask whether a node is a prefix of itself. In this
case we don’t want that. The solution to this is to drop the last
non-zero element from the coordinate. This will not prevent it from
matching against any of its ancestors, but will prevent matching
against itself. We can use the depth vector with the ↑ function to
take all but the last non-zero element. We need to remember to extend
the returned array with an extra zero column, since the shape will
be too small otherwise.

@@

We can now use Fcp and Ecp to determine which Function and Expression
nodes match for each node:

@@

We use the ⍉ function, which computes the transpose of its right
argument, to ensure that the row and column sizes match up. With the
above, we can determine the “greatest” match, which is the closest.
We can replace each 1 in the above matrices by their column numbers
(the number of the column in which that 1 occurs), and then we can
take the maximum column number in each row to determine an index in
either Fp or Ep that is the appropriate “parent” coordinate for
that node. This works because we have made sure to order the Ep and Fp
matrices lexicographically. The following shows this computation for
the expression example; we elide the function example as redundant
at this point. We also use ⌈/ instead of ⌈⌿ to reduce along
the last, rather than the first, axis.

@@

At this point we have two values, Ek and Fk, which indicate the
closest containing node that we care about for each node in the tree,
using its node coordinate.

@@

We will use these keys to perform computation over the AST and
accomplish our tasks of function lifting and expression flattening in
the next section, but we make a final note here, that we can imagine
many other operations which might be used throughout the compiler when
we care about how nodes relate to each other. The use of operators
like reduction, scan, inner and outer products allow us to obtain
the necessary information from the node coordinate matrix. Sometimes
these results might not be obvious, but the example of the “belongs
to” relationship given above demonstrates a pattern that arises
repeatedly throughout our data parallel compiler, and is therefore
particularly useful to understand.

@@

@@

If we use the key operator with Fk, we get the following (the monadic
use of ↓ converts Fc from a matrix to a vector of vectors):

@@

Notice that we have now grouped all of the relevant parts of the
tree according to which nodes would appear in their respective
functions after lifting. Refer to the original lifting example in
the introduction to verify this. Indeed. The second row in the
above example shows the internal function complete and ready to
name. Each element in the second column corresponds to the body of
one of our lifted functions. In the case of the first function, the
outer function, we have a spurious function node in the body. This
is intentional. When we lift these functions, we will replace
each spurious function node with a variable node referring to the
function’s generated name.

@@

If we use Ek as the key for the expression example, we get the
following:

@@

The J programming language [18] was the first practical,
general-purpose programming language to introduce the key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (⍤) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community [1, 17].

Fritz Henglein demonstrated a class of operations, called
discriminators, of which the Key operator is a member. [16] Namely,
a discriminator performs the same grouping computation as Key, but
does not apply a function over these groups with their keys. Henglein
provides a linear implementation of these operations.

The EigenCFA effort [22] demonstrated significant performance
improvements of a 0-CFA flow analysis by utilizing similar techniques
to those demonstrated here. In particular, encoding the AST and using
accessor functions have a very similar feel to the node coordinates and
AST encoding given here, though they have a different formulation and
spend considerable effort understanding the trade-offs of performance
associated with the different encodings, whereas the encodings here
were chosen for their clarity and directness, rather than their
performance.

Mendez-Lojo, et al. implemented a GPU version of Inclusion-based
Points-to Analysis [20] that also focuses on adapting data structures
and algorithms to efficiently execute on the GPU. In particular,
they use similar techniques of prefix sums and sorts to achieve some
of their adaptation to the GPU, Additionally, they have clever and
efficient methods of representing graphs on the GPU which enable
dynamic rewriting of the graph.

The APEX compiler [2] developed vectorized approaches to handling
certain analyses to compile traditional APL, including a SIMD tokenizer
[4]. It uses a SSA representation, and converts the dynamic scope of
traditional APL functions into a static form early on. It also uses
a matrix format to represent the AST. Traditional APL did not have
nested function definitions, however, and thus the APEX compiler does
not have any specific approaches to dealing with function lifting.

Bernecky further identified methods of reducing or optimizing the
computational complexity or cost of certain array operations, allowing
improved performance of easy to understand array expressions [3].

Timothy Budd implemented a compiler [5, 6] for APL which targeted
vector processors as well as one for C code. They used a method of
lazy evaluation to avoid intermediate data copying. Budd provided
thoughts and some ideas on how the compiler might be implemented in
parallel as well.

Walter Schwarz implemented an APL to C compiler for the ACORN system
targeting the CM-2 machine, demonstrating performance potential for
APL as a massively parallel language [21].

W. Ching and D. Ju have spent significant work on the ELI language
and other APL-class language
 implementations, especially on parallelized code and optimization. [8,
 9, 10, 11, 13, 14, 15]

J. D. Bunda and J. A. Gerth presented a method for doing table driven
parsing of APL which suggested a parallel optimization for parsing,
but did not elucidate the algorithm [7].

@@

% Bibliography goes here

>>>>

There has been some interesting use of GPUs to accelerate certain
compiler analyses and passes.  Unfortunately, attempts to parallelize
compilers usually takes the form of a great deal of work spent
to highly optimize a single compiler pass. The results are often
fragile and require a great deal of effort to maintain and keep in
line with the rest of the compiler. This integration problem makes it
difficult to scale the ``ad hoc'' parallelization effort of individual
compiler passes to a complete system. This is particularly true in
situations where the compiler passes may change rapidly, such as with
XSLT or document publishing, where it is very common for the document
preparer to write or tweak transformation passes to suit the specific
class of documents being handled. In these cases, the expertise and
time involved in parallelizing such transformations makes the ad hoc
approach mostly intractable.

Furthermore, compilers are increasingly becoming a part of a ``live''
and interactive feedback loop.  Just in time compilation and on
the fly static compilation are both used to provide custom binaries
tuned to specific tasks on the fly. These techniques are used in web
services and large scale databases, where in-memory databases may
construct functions on the fly to execute. There is great interest
in moving such systems on to the GPU or other highly data-parallel
architectures, but it would significantly hamper the performance of
the system if the compilation phase of query execution were required
to occur on the CPU, especially at scale where there may be many more
GPU accelerators than CPU resources.

@@

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{∧}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives}
\label{tab:scalarprims}
\end{table}

@@

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Monadic/Unary Operators, each takes a single left
operand and describes a function operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a function along cells of an array \\
\end{tabular}
\caption{Primitive Dyadic/Binary Operators, each takes a left
 and right operand and describes a function operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

@@

@@

We represent the AST as a matrix of 3 columns and n rows, one for
each node in the tree. The first column contains the inter-node
relationships in the form of a depth vector. The second column is a
vector of the node types, while the third contains the ``value''
of the node, such as the name in a variable.

@@

The depth vector for these trees we name \verb;Fd; and \verb;Ed;, respectively:

@@

The node types we call \verb;Ft; and \verb;Et;:

@@

And finally, we call the values vectors \verb;Fv; and \verb;Ev;:

@@

We combine these to form the respective AST matrices. We write \verb;A,B;
to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
vector into a 1-column matrix. Thus, the two tree matrices are given
by the following expressions:

\begin{verbatim}
      Fd,Ft,⍪Fv
0 F f
1 E 0
2 F 0
3 E 0
4 V ⍵
4 P +
4 V ⍵
2 A 0
3 N 7
      Ed,Et,⍪Ev
0 E v
1 E 0
2 V a
2 P +
2 V b
1 P ÷
1 E 0
2 V c
2 P ×
2 V d
\end{verbatim}

@@

\begin{enumerate}[noitemsep]
\item Are the nodes the same?
\item Are they siblings?
\item Does one appear ``earlier'' in the tree?
\item Are they at the same depth?
\item Is one an ancestor of another?
\end{enumerate}

@@

We write \verb;f⌿A; to reduce the first axis of \verb;A; using function \verb;f;. Thus,
\verb;+⌿V; is the sum of the elements in vector \verb;V;. The function \verb;x⌈y;
gives the maximum of its two arguments. So, the depth of each tree
is given by the following:

@@

The function table or outer product of \verb;f; over vectors \verb;U; and \verb;V; is
written \verb;U ∘.f V; giving a \verb;U V; shaped matrix as a result. Thus,
\verb;(⍳3)∘.×⍳3; gives a small multiplication table:

@@

If we use \verb;=; instead, we have a Boolean identity matrix:

@@

If we use \verb;∘.=; on the depth vector and its set of depths instead,
we see an expanded Boolean representation of the depth vector:

@@

These matrices let us see the nesting features of each tree more
visually, but also suggest another step. We can compute a sum scan
with \verb;+⍀;, also called a prefix sum, along the first axis. Applying
this function on the above matrices leads to an interesting result:

@@

The expression \verb;V f⍤¯1⊢M; applies \verb;f; to corresponding elements of
\verb;V; and rows of \verb;M;:

@@

If \verb;n↑V; takes the first \verb;n; elements of \verb;V;, then we can obtain coordinate
matrices from the prefix sums by noting that the spurious digits
all come after column \verb;d+1; where \verb;d; is the depth corresponding to that
coordinate. The following gives a complete expression for computing
a node coordinate matrix from a depth vector, shown using \verb;Fd; and \verb;Fe;:

@@

The simplest operation over a node coordinate is to extract the depth
of the node. The expression \verb;C⍳0; finds the first occurrence of \verb;0; in \verb;C;
and returns the index of that occurrence. (The application of ⍳ here
is its dyadic form, meaning that it receives two arguments. Previous
uses of ⍳ have used its monadic form, which generates all indices,
instead of finding a specific index. Most APL functions have a
dyadic and monadic form.)  Thus, the depth of a node is given by
the following:

@@

We write \verb;(f g h); to represent the composition of functions \verb;f;, \verb;g;,
and \verb;h; as a function train. This is used with another shorthand of
composition according to the following equivalences:

\begin{verbatim}
A(f g h)B ←→ (A f B) g (A h B)
A(0 f h)B ←→ 0 f (A h B) ⍝ Constant Case
A(f g)B   ←→ f (A g B)
\end{verbatim}

@@

Here \verb;P; is the coordinate we wish to check against \verb;C;, to determine
whether \verb;C; is a descendant. The functions are all standard logical
functions extended to apply point-wise over arrays. The function
\verb;⊢; always returns its right argument. The function \verb;(=∨0=⊢);
reads intuitively as, ``equal or right element is zero.'' In this
case to make this a predicate we combine this with the ``for all''
reduction using \verb;∧⌿;. This particular pattern is a special case
of inner product, which we can compute using \verb;f.g;. Thus, \verb;+.×; is a
function that computes traditional matrix-matrix multiplication when
applied to two matrices. However, we can also use it to compute the
above reduction and application:

@@

This extends the reduction to allow us to use matrix values for
\verb;C; and \verb;P; instead of vectors, and thus determine whether rows in \verb;C;
are descendants of multiple candidates given by \verb;P;. This becomes
important to using the prefix function. In our examples, we want
to determine the nearest ancestor of a certain node type to which a
node belongs. In the case of function lifting, we want to determine
the nearest ancestor function, in the case of expression flattening,
we want to determine the nearest ancestor expression node. We can
extract the rows we care about for each example using \verb;⌷;, where \verb;i⌷M;
returns the i-th row of \verb;M;. In the following expressions, we use \verb;0 2;
rather than \verb;¯1; in \verb;⍤; to indicate that \verb;⌷; will receive the entire
matrix on the right for each scalar index on the left.

@@

The matrices \verb;Fp; and \verb;Ep; correspond to the function and expression
node coordinates in our two examples, respectively. We could now
use \verb;Fp; and \verb;Ep; to compare against \verb;Fc; and \verb;Ec; and determine which nodes
contained which nodes in the tree. However, our prefix function will
return 1 when we ask whether a node is a prefix of itself. In this
case we don’t want that. The solution to this is to drop the last
non-zero element from the coordinate. This will not prevent it from
matching against any of its ancestors, but will prevent matching
against itself. We can use the depth vector with the \verb;↑; function to
take all but the last non-zero element. We need to remember to extend
the returned array with an extra zero column, since the shape will
be too small otherwise.

@@

We can now use \verb;Fcp; and \verb;Ecp; to determine which Function and Expression
nodes match for each node:

@@

We use the ⍉ function, which computes the transpose of its right
argument, to ensure that the row and column sizes match up. With the
above, we can determine the ``greatest'' match, which is the closest.
We can replace each 1 in the above matrices by their column numbers
(the number of the column in which that 1 occurs), and then we can
take the maximum column number in each row to determine an index in
either \verb;Fp; or \verb;Ep; that is the appropriate ``parent'' coordinate for
that node. This works because we have made sure to order the \verb;Ep; and \verb;Fp;
matrices lexicographically. The following shows this computation for
the expression example; we elide the function example as redundant
at this point. We also use \verb;⌈/; instead of \verb;⌈⌿; to reduce along
the last, rather than the first, axis.

@@

At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
closest containing node that we care about for each node in the tree,
using its node coordinate.

@@

We will use these keys to perform computation over the AST and
accomplish our tasks of function lifting and expression flattening in
the next section, but we make a final note here, that we can imagine
many other operations which might be used throughout the compiler when
we care about how nodes relate to each other. The use of operators
like reduction, scan, inner and outer products allow us to obtain
the necessary information from the node coordinate matrix. Sometimes
these results might not be obvious, but the example of the ``belongs
to'' relationship given above demonstrates a pattern that arises
repeatedly throughout our data parallel compiler, and is therefore
particularly useful to understand.

@@

The Key operator (written \verb;⌸;), takes a single function
that expects two arguments and returns a function which takes two
arguments. The left argument is a set of keys, and the right argument
is a set of corresponding elements associated with those keys. In our
case, we provide matrices for both of these arguments, so each pair
of rows corresponds to a key-value pair. A simple example of the key
operation at work is to compute a histogram (\verb;≢; here can be thought
of as ``tally'' or count):

To understand a bit better how the Key operator applies its function,
consider the function \verb;{⍺ ⍵}; which returns the pair of its right
and left arguments. If we apply it to the same value as above, we
get the following:

In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
corresponding AST. We also will drop off the first row in each AST
using \verb;1↓; since this node ``contains'' everything. In a complete
AST this is usually the Module boundary node which contains the entire
set of functions and values in the module.

@@

If we use the key operator with \verb;Fk;, we get the following (the monadic
use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):

@@

Notice that we have now grouped all of the relevant parts of the
tree according to which nodes would appear in their respective
functions after lifting. Refer to the original lifting example in
the introduction to verify this. Indeed, the second row in the
above example shows the internal function complete and ready to
name. Each element in the second column corresponds to the body of
one of our lifted functions. In the case of the first function, the
outer function, we have a spurious function node in the body. This
is intentional. When we lift these functions, we will replace
each spurious function node with a variable node referring to the
function’s generated name.

@@

If we use \verb;Ek; as the key for the expression example, we get the
following:

@@

The J programming language \cite{hui2014key} was the first practical,
general-purpose programming language to introduce the key operator as a
primitive operator with the presumption of its general usefulness. The
rank operator (⍤) used throughout this treatment also derives from
the J traditional, receiving particular interest throughout the APL
community \cite{bernecky1987rank,hui1995rank}.

Fritz Henglein demonstrated a class of operations, called
discriminators, of which the Key operator is a member.  
\cite{henglein2013dd}
Namely,
a discriminator performs the same grouping computation as Key, but
does not apply a function over these groups with their keys. Henglein
provides a linear implementation of these operations.

The EigenCFA effort \cite{prabhu2011eigencfa} demonstrated significant performance
improvements of a 0-CFA flow analysis by utilizing similar techniques
to those demonstrated here. In particular, encoding the AST and using
accessor functions have a very similar feel to the node coordinates and
AST encoding given here, though they have a different formulation and
spend considerable effort understanding the trade-offs of performance
associated with the different encodings, whereas the encodings here
were chosen for their clarity and directness, rather than their
performance.

Mendez-Lojo, et al. implemented a GPU version of Inclusion-based
Points-to Analysis \cite{mendez2012inclusion} that also focuses on 
adapting data structures
and algorithms to efficiently execute on the GPU. In particular,
they use similar techniques of prefix sums and sorts to achieve some
of their adaptation to the GPU, Additionally, they have clever and
efficient methods of representing graphs on the GPU which enable
dynamic rewriting of the graph.

The APEX compiler \cite{bernecky1997apex} developed vectorized approaches to handling
certain analyses to compile traditional APL, including a SIMD tokenizer
\cite{bernecky2003tokenizer}.
 It uses a SSA representation, and converts the dynamic scope of
traditional APL functions into a static form early on. It also uses
a matrix format to represent the AST. Traditional APL did not have
nested function definitions, however, and thus the APEX compiler does
not have any specific approaches to dealing with function lifting.

Bernecky further identified methods of reducing or optimizing the
computational complexity or cost of certain array operations, allowing
improved performance of easy to understand array expressions
\cite{bernecky1999reducing}.

Timothy Budd implemented a compiler \cite{budd1984apl,budd2012apl} for APL which targeted
vector processors as well as one for C code. They used a method of
lazy evaluation to avoid intermediate data copying. Budd provided
thoughts and some ideas on how the compiler might be implemented in
parallel as well.

Walter Schwarz implemented an APL to C compiler for the ACORN system
targeting the CM-2 machine, demonstrating performance potential for
APL as a massively parallel language \cite{schwarz1991acorn}.

W. Ching and D. Ju have spent significant work on the ELI language
and other APL-class language
 implementations, especially on parallelized code and optimization. 
\cite{ching2000design,ching1994experimental,ching1993primitive,
ching1990automatic,hendriks1990sparse,ju1991exploitation,
ju1991performance}

J. D. Bunda and J. A. Gerth presented a method for doing table driven
parsing of APL which suggested a parallel optimization for parsing,
but did not elucidate the algorithm \cite{bunda1984apl}.

@@

\bibliographystyle{plain}
\bibliography{biblio}
